---
title: "Cleaning Bad Longitude in HURDAT"
author: "Tim Trice"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cleaning Bad Longitude in HURDAT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      fig.width = 7, 
                      fig.height = 5)
library(data.table)
library(dplyr)
library(ggplot2)
library(Hurricanes)
library(knitr)
library(maps)
library(readr)
```

There is a known issue in the raw file for the Atlantic basin where longitude values are incorrect. This file demonstrates cleaning that data and correcting the necessary data tables.

We start by getting our raw datasets with `raw_hurdat()`.

```{r}
raw_hurdat()
```

I'll bind `atlantic` and `nepac` for easier analysis.

```{r}
df <- bind_rows(nepac, atlantic)
```

Clean latitude and longitude

```{r}
df$Latitude <- clean_lat(df$Latitude)
df$Longitude <- clean_lon(df$Longitude)
```

```{r}
summary(df$Longitude)
```

There is at least one value of -359.1&deg;W. We should not have any longitude values $\pm$ 180&deg; as our dataset only covers the Atlantic, north eastern and central Pacific; effectively the northwestern hemisphere. 

To get a better idea of where our longitude values stand, we'll create a plot.

```{r}
world_map <- map_data("world")

bp <- list(geom_polygon(data = world_map, aes(x = long, y = lat, group = group), 
                 colour = "grey10", fill = "white"), 
           geom_point(size = 0.5, colour = "red"))

ggplot(df, aes(x = Longitude, y = Latitude)) + 
    bp
```

Our latitude values seem fine - we're expecting and received only positive values.

But we have some bad longitude values off to the left that's blowing the scale of our map. So, let's filter our data frame on this.

```{r}
tmp <- df %>% 
 filter(Longitude < -180)
```

What storms are we dealing with here?

```{r}
unique(tmp$Key)
```

Our `Key` value is formatted like "AABBCCCC" where AA is a Basin identifier, BB is the number of the storm for the specific year and CCCC is the specific year. So, `AA == AL` is for Atlantic storms and is in our `atlantic` dataset.

Is this error an issue with our `Hurricanes` library? What did our original data look like:

```{r}
tmp_atlantic <- filter(atlantic, Key == "AL051952")
print(tmp_atlantic$Longitude)
```

So, we have two bad values at the end - "359.1W" and "358.4W". Let's verify this was in our original dataset. 

```{r}
raw_atlantic <- read_csv(
    "http://www.nhc.noaa.gov/data/hurdat/hurdat2-1851-2015-021716.txt", 
    skip = 23823L, n_max = 2, col_names = FALSE)
raw_atlantic
```

We notice in `X6` the bad values. So, this isn't the `Hurricanes` package erring when parsing the data; it's just bad data. But we also notice a pattern with the longitude values. Let's look at them again:

```{r}
print(tmp_atlantic$Longitude)
```

Our values are descending - the storm is moving eastward - until it gets to 0&deg; where they should start *increasing* from 0. In other words, our 359.1W value should actually be +0.9 (360 - 359.1) and 358.4 should be +1.6. 

It's important to note when we run `clean_lat` and `clean_lon`, values in the northern and eastern hemisphere are positive while values in the southern and western hemisphere are negative. That our storms *originated* in the northwestern hemisphere means all of our latitude values should be positive and our longitude values should be negative. We already verified `Latitude` seems fine. 

But now we see that we had storms that actually crossed hemispheres. So, now we know as far as `Longitude` is concerned that we should have both positive and negative values. 

In our original plot of `Longitude` we saw we had two records less than -300&deg;. `clean_lat` takes `Latitude` values with the "W" hemisphere identifider and simply multiplies them by -1 making them negative, as they should be. 

But, because our raw data contained "359.1W" and "358.4W", we got results of -359.1 and -358.4 - the two outliers on the left-side of our chart. 

$$
Lon = 
\begin{cases}
360 + Lon, & Lon < -180 \\
Lon, & Lon
\end{cases}
$$

So, by adding 360 to any `Longitude` value < -180, we get a positive `Longitude` value which corrects our two observations. 

```{r}
df$Longitude <- sapply(df$Longitude, function(x) { ifelse(x < -180, 360 + x, x)})
```

And now our plot again.

```{r}
ggplot(df, aes(x = Longitude, y = Latitude)) + 
    bp
```

This is as expected and tells us our data at least *looks* good. We're done.
